---
title: "Your First R Session"
teaching: 60
exercises: 30
---

:::::::::::::::::::::::::::::::::::::: questions

- How does RStudio compare to the SPSS interface?
- How do I import data, including SPSS `.sav` files?
- How do I get descriptive statistics and frequency tables in R?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Navigate the RStudio interface and identify the equivalent of SPSS panels
- Import CSV and SPSS `.sav` files into R
- Run basic descriptive statistics and frequency tables
- Inspect variables and data structure

::::::::::::::::::::::::::::::::::::::::::::::::

## RStudio orientation

When you open RStudio for the first time, you see four panes. If you have used
SPSS before, each one has a rough equivalent:

| RStudio pane | Location | SPSS equivalent | What it does |
|---|---|---|---|
| **Source Editor** | Top-left | Syntax Editor | Where you write and save your code (scripts) |
| **Console** | Bottom-left | Output Viewer | Where R runs commands and prints results |
| **Environment** | Top-right | Data View header | Lists all objects (datasets, values) currently in memory |
| **Files / Plots / Help** | Bottom-right | *(no equivalent)* | File browser, plot preview, and built-in documentation |

The key difference from SPSS: in SPSS, you usually have *one* dataset open at
a time and interact through menus. In RStudio, you write instructions in the
Source Editor (top-left), send them to the Console (bottom-left), and the
results appear either in the Console or the Plots pane.

::::::::::::::::::::::::::::::::::::: callout

## The Source Editor is your new best friend

In SPSS, many users never open the Syntax Editor — they click menus instead.
In R, the Source Editor *is* how you work. Think of it as a recipe: you write
the steps once, and you (or anyone else) can re-run them at any time.

Save your scripts with the `.R` extension. You will thank yourself later.

::::::::::::::::::::::::::::::::::::::::::::::::

## Objects and assignment

In SPSS, when you compute a new variable, it appears as a column in your
dataset. In R, everything you create is stored as a named **object**.

You create objects with the **assignment operator** `<-` (a less-than sign
followed by a hyphen). Read it as "gets" or "is assigned".

```{r}
# Store a number
population <- 106739

# Store text (called a "character string" in R)
island <- "Aruba"

# Store the result of a calculation
density <- population / 180  # Aruba is about 180 km²
```

To see the value of an object, type its name and run it:

```{r}
population
island
density
```

::::::::::::::::::::::::::::::::::::: callout

## Why `<-` and not `=`?

You will see some people use `=` for assignment, and it works in most cases.
However, the R community convention is `<-`. It makes your code easier to read
because `=` is also used inside function arguments (as you will see shortly).

In RStudio, the keyboard shortcut **Alt + -** (Alt and the minus key) types
`<-` for you automatically.

::::::::::::::::::::::::::::::::::::::::::::::::

## Functions: R's version of menu clicks

In SPSS, you click **Analyze > Descriptive Statistics > Descriptives** and a
dialog box appears. In R, you call a **function** instead. A function has a
name, and you pass it **arguments** inside parentheses.

```{r}
# round() is a function. 592.777 is the input, digits = 1 is an option.
round(592.777, digits = 1)
```

```{r}
# sqrt() calculates a square root
sqrt(density)
```

The pattern is always: `function_name(argument1, argument2, ...)`. This is
the R equivalent of filling in an SPSS dialog box — the function name is the
menu item, and the arguments are the fields you would fill in.

## Packages: extending R

R comes with many built-in functions, but its real power comes from
**packages** — add-on libraries written by other users. Think of them as SPSS
modules, except they are free.

There are two steps:

1. **Install** the package (once per computer, like installing an app):

```{r, eval=FALSE}
install.packages("tidyverse")
```

2. **Load** the package (once per session, like opening an app):

```{r}
library(tidyverse)
```

::::::::::::::::::::::::::::::::::::: callout

## `install.packages()` vs `library()`

A common source of confusion for beginners:

- `install.packages("tidyverse")` — downloads and installs the package. You
  only need to do this **once** (or when you want to update). Note the
  **quotation marks**.
- `library(tidyverse)` — loads a package that is already installed so you can
  use it in your current session. You do this **every time** you start R. No
  quotation marks needed (though they work too).

Analogy: `install.packages()` is buying a book and putting it on your shelf.
`library()` is taking the book off the shelf and opening it.

::::::::::::::::::::::::::::::::::::::::::::::::

## Importing data

### CSV files with `read_csv()`

The most common data format in R is CSV (comma-separated values). The
`readr` package (loaded as part of `tidyverse`) provides `read_csv()`:

```{r}
visitors <- read_csv("data/aruba_visitors.csv")
```

R prints a summary of the column types it detected. This is equivalent to
opening a CSV in SPSS via **File > Open > Data** and checking the variable
types.

### SPSS `.sav` files with `haven`

If you have existing SPSS datasets, the `haven` package reads them directly —
including variable labels and value labels:

```{r, eval=FALSE}
# install.packages("haven")  # run once if needed
library(haven)
spss_data <- read_sav("path/to/your/file.sav")
```

This means you do not have to convert your SPSS files to CSV first. R reads
them as-is.

## Exploring your data

Now that we have the `visitors` dataset loaded, let us explore it. Each of the
functions below is the R equivalent of something you would do in SPSS.

### `View()` — the Data View equivalent

```{r, eval=FALSE}
View(visitors)
```

This opens a spreadsheet-like viewer in RStudio, just like SPSS Data View. You
can scroll, sort columns by clicking headers, and filter. (Note the capital V.)

### `head()` — see the first few rows

```{r}
head(visitors)
```

This is faster than `View()` when you just want a quick look. By default it
shows 6 rows. You can change that: `head(visitors, n = 10)`.

### `str()` — the Variable View equivalent

In SPSS, you would switch to **Variable View** to see variable names, types,
and labels. In R, `str()` does the same thing:

```{r}
str(visitors)
```

This tells you: how many observations (rows), how many variables (columns),
and the type of each variable (`num` for numbers, `chr` for text).

### `glimpse()` — a tidyverse alternative to `str()`

The `glimpse()` function from `dplyr` gives similar information in a tidier
format:

```{r}
glimpse(visitors)
```

### `summary()` — Descriptives in one command

In SPSS: **Analyze > Descriptive Statistics > Descriptives**. In R:

```{r}
summary(visitors)
```

For numeric columns, you get the minimum, maximum, mean, median, and
quartiles. For character columns, you get the length and type.

### `table()` — Frequency tables

In SPSS: **Analyze > Descriptive Statistics > Frequencies**. In R:

```{r}
table(visitors$origin)
```

The `$` operator extracts a single column from a data frame. So
`visitors$origin` means "the origin column from the visitors dataset" — like
clicking on a single variable in SPSS.

You can also make two-way frequency tables:

```{r}
table(visitors$year, visitors$origin)
```

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

## Pacing notes

- Spend time on the RStudio pane orientation. Have participants identify each
  pane on their own screen before moving on.
- The `<-` assignment operator trips people up. Give them a few minutes to
  practice creating objects with different names and values.
- When loading tidyverse, the startup messages can be alarming to beginners.
  Reassure them that the "Attaching packages" and "Conflicts" messages are
  normal and expected.
- For `read_csv()`, make sure everyone has the data file in the right place.
  This is the most common point of failure. Walk around and check.

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Explore the Aruba visitors dataset

Import the Aruba visitors dataset and answer the following questions using R
functions. Write your code in the Source Editor and run each line.

1. How many rows and how many columns does the dataset have?
2. What data type is the `origin` column? What data type is `visitors_stayover`?
3. What is the mean `avg_spending_usd` across all rows?
4. How many rows are there for each origin country?

:::::::::::::::::::::::: solution

## Solution

```{r}
# Load the data (if not already loaded)
library(tidyverse)
visitors <- read_csv("data/aruba_visitors.csv")
```

**Question 1:** How many rows and columns?

```{r}
# Either of these works:
dim(visitors)
glimpse(visitors)
```

The dataset has 120 rows and 9 columns (5 years x 4 quarters x 6 origin
countries = 120 rows).

**Question 2:** Data types?

```{r}
str(visitors)
```

`origin` is character (`chr`), `visitors_stayover` is numeric (`num`).

**Question 3:** Mean average spending?

```{r}
summary(visitors$avg_spending_usd)
```

The mean `avg_spending_usd` is shown in the summary output. You can also get
just the mean with:

```{r}
mean(visitors$avg_spending_usd)
```

**Question 4:** Rows per origin country?

```{r}
table(visitors$origin)
```

Each origin country has 20 rows (4 quarters x 5 years).

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Practice with objects and functions

1. Create an object called `my_island` that stores the text `"Aruba"`.
2. Create an object called `area_km2` that stores the value `180`.
3. Use the `nchar()` function to count the number of characters in `my_island`.
4. Use `round()` to round the mean of `avg_spending_usd` to the nearest whole
   number. (Hint: you can put one function inside another.)

:::::::::::::::::::::::: solution

## Solution

```{r}
# 1 and 2: Create objects
my_island <- "Aruba"
area_km2 <- 180

# 3: Count characters
nchar(my_island)

# 4: Round the mean spending
round(mean(visitors$avg_spending_usd), digits = 0)
```

Nesting functions (putting one inside another) is common in R. R evaluates
from the inside out: first it calculates `mean(visitors$avg_spending_usd)`,
then it passes that result to `round()`.

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Summary

You have now completed your first hands-on R session. You can:

- Find your way around RStudio
- Create objects and use functions
- Install and load packages
- Import a CSV file
- Inspect your data with `View()`, `head()`, `str()`, `glimpse()`, `summary()`,
  and `table()`

In SPSS terms, you have learned the equivalent of opening a dataset, switching
between Data View and Variable View, and running Descriptives and Frequencies.
The difference is that everything you did is saved in a script that you can
re-run at any time.

::::::::::::::::::::::::::::::::::::: keypoints

- RStudio is your workspace — it combines a script editor, console, and data viewer
- `haven::read_sav()` imports SPSS files directly, preserving labels
- `summary()`, `table()`, and `str()` replace the Descriptives and Frequencies menus in SPSS

::::::::::::::::::::::::::::::::::::::::::::::::
