---
title: "Statistical Analysis in R"
teaching: 80
exercises: 40
---

:::::::::::::::::::::::::::::::::::::: questions

- How do I run t-tests, ANOVA, correlations, and regression in R?
- How does R output compare to SPSS output tables?
- How do I extract and report results?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Run independent and paired samples t-tests
- Conduct one-way ANOVA with post-hoc tests
- Calculate correlations and run simple linear regression
- Interpret R output by mapping it to familiar SPSS output tables

::::::::::::::::::::::::::::::::::::::::::::::::

## From SPSS dialogs to R functions

In SPSS, every statistical test lives behind a menu: **Analyze > Compare Means**,
**Analyze > Correlate**, and so on. In R, each test is a single function call. The
table below maps the SPSS dialogs you already know to their R equivalents:

| Analysis                     | SPSS menu path                                         | R function                                  |
|------------------------------|--------------------------------------------------------|---------------------------------------------|
| Independent-samples t-test   | Analyze > Compare Means > Independent-Samples T Test   | `t.test(y ~ group, data = df)`              |
| Paired-samples t-test        | Analyze > Compare Means > Paired-Samples T Test        | `t.test(x, y, paired = TRUE)`               |
| One-way ANOVA                | Analyze > Compare Means > One-Way ANOVA                | `aov(y ~ group, data = df)` + `summary()`   |
| Post-hoc tests               | (checkbox in ANOVA dialog)                             | `TukeyHSD()`                                |
| Bivariate correlation        | Analyze > Correlate > Bivariate                        | `cor.test(df$x, df$y)`                      |
| Linear regression            | Analyze > Regression > Linear                          | `lm(y ~ x1 + x2, data = df)` + `summary()` |

Let's load our data and packages:

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)

visitors <- read_csv("data/aruba_visitors.csv")
```

### T-tests

#### Independent-samples t-test

In SPSS, you would go to **Analyze > Compare Means > Independent-Samples T
Test**, move your test variable to the "Test Variable(s)" box, move your grouping
variable to the "Grouping Variable" box, and define the two groups.

In R, it is one line. Let's test whether the average spending differs between
US and Netherlands visitors:

```{r t-test-data}
# Filter to just the two groups we want to compare
us_nl <- visitors |>
  filter(origin %in% c("United States", "Netherlands"))

# Run the independent-samples t-test
t_result <- t.test(avg_spending_usd ~ origin, data = us_nl)
t_result
```

::::::::::::::::::::::::::::::::::::: callout

## Reading the t-test output --- SPSS comparison

The R output gives you the same information as the SPSS "Independent Samples
Test" table, just arranged differently:

| SPSS output column          | R output line                           |
|-----------------------------|-----------------------------------------|
| t                           | `t = ...`                               |
| df                          | `df = ...`                              |
| Sig. (2-tailed)             | `p-value = ...`                         |
| Mean Difference             | difference in means shown in estimates  |
| 95% CI of the Difference    | `95 percent confidence interval:`       |

The key difference: SPSS shows Levene's test for equality of variances
automatically. R's `t.test()` uses the Welch correction by default (which does
**not** assume equal variances). This is actually the better default --- many
statisticians recommend always using the Welch t-test.

If you need the equal-variances version (the "Equal variances assumed" row in
SPSS), add `var.equal = TRUE`:

```r
t.test(avg_spending_usd ~ origin, data = us_nl, var.equal = TRUE)
```

::::::::::::::::::::::::::::::::::::::::::::::::

#### Paired-samples t-test

A paired t-test compares two measurements on the same cases. Let's compare Q1
vs. Q3 hotel occupancy rates (the same hotels measured in different quarters):

```{r paired-t-test}
# Get Q1 and Q3 data for each year
q1_data <- visitors |>
  filter(quarter == "Q1") |>
  arrange(year, origin) |>
  pull(hotel_occupancy_pct)

q3_data <- visitors |>
  filter(quarter == "Q3") |>
  arrange(year, origin) |>
  pull(hotel_occupancy_pct)

# Paired t-test
t.test(q1_data, q3_data, paired = TRUE)
```

In SPSS this would be **Analyze > Compare Means > Paired-Samples T Test**,
where you select the two variables as a pair.

### ANOVA

#### One-way ANOVA

In SPSS: **Analyze > Compare Means > One-Way ANOVA**. Move the dependent
variable to the "Dependent List" and the factor to "Factor".

Let's test whether satisfaction scores differ across origin countries:

```{r anova}
# Fit the ANOVA model
anova_model <- aov(satisfaction_score ~ origin, data = visitors)

# View the ANOVA table
summary(anova_model)
```

::::::::::::::::::::::::::::::::::::: callout

## Reading the ANOVA table --- SPSS comparison

| SPSS output column | R output column |
|--------------------|-----------------|
| Sum of Squares     | `Sum Sq`        |
| df                 | `Df`            |
| Mean Square        | `Mean Sq`       |
| F                  | `F value`       |
| Sig.               | `Pr(>F)`        |

The layout is almost identical --- R just uses slightly different column names.

::::::::::::::::::::::::::::::::::::::::::::::::

#### Post-hoc tests with TukeyHSD

If the ANOVA is significant, you want to know *which* groups differ. In SPSS,
you check the "Post Hoc" button in the ANOVA dialog and select Tukey. In R:

```{r tukey}
TukeyHSD(anova_model)
```

This gives you pairwise comparisons with adjusted p-values, the difference in
means, and 95% confidence intervals --- the same information as the SPSS
"Multiple Comparisons" table.

You can also visualize the post-hoc results:

```{r tukey-plot}
plot(TukeyHSD(anova_model), las = 1, cex.axis = 0.7)
```

### Correlation

In SPSS: **Analyze > Correlate > Bivariate**. Move variables to the "Variables"
box and select Pearson, Spearman, or both.

Let's test the correlation between average spending and satisfaction:

```{r correlation}
cor.test(visitors$avg_spending_usd, visitors$satisfaction_score)
```

::::::::::::::::::::::::::::::::::::: callout

## Reading correlation output --- SPSS comparison

| SPSS output                  | R output                         |
|------------------------------|----------------------------------|
| Pearson Correlation          | `cor` (the estimate at the end)  |
| Sig. (2-tailed)              | `p-value`                        |
| N                            | shown in the data, not in output |
| 95% CI                       | `95 percent confidence interval` |

One advantage of R: `cor.test()` gives you a confidence interval for the
correlation by default. SPSS does not show this unless you use syntax.

::::::::::::::::::::::::::::::::::::::::::::::::

For a correlation matrix of multiple variables (like the SPSS correlation table),
use `cor()`:

```{r cor-matrix}
visitors |>
  select(visitors_stayover, avg_stay_nights, avg_spending_usd,
         hotel_occupancy_pct, satisfaction_score) |>
  cor(use = "complete.obs") |>
  round(3)
```

### Linear regression

In SPSS: **Analyze > Regression > Linear**. Move the dependent variable to
"Dependent" and independent variables to "Independent(s)".

Let's predict average spending from stay nights and origin country:

```{r regression}
reg_model <- lm(avg_spending_usd ~ avg_stay_nights + origin, data = visitors)
summary(reg_model)
```

::::::::::::::::::::::::::::::::::::: callout

## Reading regression output --- SPSS comparison

The `summary()` output contains everything from the SPSS regression output
tables, but in a more compact format:

| SPSS table            | R output section                     |
|-----------------------|--------------------------------------|
| Model Summary (R-sq)  | `Multiple R-squared`, `Adjusted R-squared` at the bottom |
| ANOVA table (F-test)  | `F-statistic` at the very bottom     |
| Coefficients table    | The `Coefficients:` section          |
| B (unstandardized)    | `Estimate` column                    |
| Std. Error            | `Std. Error` column                  |
| t                     | `t value` column                     |
| Sig.                  | `Pr(>|t|)` column                    |

Note: R does **not** give you standardized coefficients (Beta) by default. To
get those, scale your variables first with `scale()`, or use the `lm.beta`
package.

::::::::::::::::::::::::::::::::::::::::::::::::

## Reading R output with `broom::tidy()`

The raw R output is fine for interactive exploration, but it is hard to export
or combine with other results. The `broom` package converts statistical output
into tidy data frames --- one row per term, columns for estimate, standard
error, test statistic, and p-value.

```{r broom-tidy}
# Tidy the t-test result
tidy(t_result)
```

```{r broom-regression}
# Tidy the regression coefficients
tidy(reg_model)
```

```{r broom-glance}
# Get model-level statistics (R-squared, F, etc.)
glance(reg_model)
```

The `tidy()` output is a regular data frame, which means you can filter it,
arrange it, or export it to CSV --- something that is surprisingly difficult
with SPSS output. This is extremely useful for reporting: you can extract
exactly the numbers you need and drop them into a table or a plot.

:::::::::::::::::::::::::::::::::::::::::::: instructor

## Instructor note

This is a good time to reinforce the reproducibility advantage. In SPSS, if a
reviewer asks you to re-run an analysis with a different subset, you have to
click through the dialogs again. In R, you change one line of code and re-run
the script.

Emphasize that `broom::tidy()` produces a data frame --- this means students
can use all the dplyr verbs they learned in Episode 3 on their statistical
results (filtering significant results, arranging by p-value, etc.).

::::::::::::::::::::::::::::::::::::::::::::::::::::::::

## A complete analysis workflow

Let's put everything together in a workflow that mirrors what you would do in
SPSS, but entirely in code. Our research question: **Does average spending
differ by origin country, and what predicts higher spending?**

```{r workflow-step1}
# Step 1: Descriptive statistics by group
visitors |>
  group_by(origin) |>
  summarise(
    n = n(),
    mean_spending = mean(avg_spending_usd),
    sd_spending = sd(avg_spending_usd),
    mean_satisfaction = mean(satisfaction_score)
  ) |>
  arrange(desc(mean_spending))
```

```{r workflow-step2}
# Step 2: Visualize the distribution
ggplot(visitors, aes(x = reorder(origin, avg_spending_usd, FUN = median),
                     y = avg_spending_usd)) +
  geom_boxplot(fill = "#2a9d8f", alpha = 0.6) +
  labs(title = "Average Spending by Origin Country",
       x = "Country of Origin",
       y = "Average Spending (USD)") +
  theme_minimal()
```

```{r workflow-step3}
# Step 3: ANOVA --- is there a significant difference?
spending_anova <- aov(avg_spending_usd ~ origin, data = visitors)
summary(spending_anova)
```

```{r workflow-step4}
# Step 4: Post-hoc --- which groups differ?
tukey_results <- TukeyHSD(spending_anova)
tidy(tukey_results) |>
  filter(adj.p.value < 0.05) |>
  arrange(adj.p.value)
```

```{r workflow-step5}
# Step 5: Regression --- what predicts spending?
spending_model <- lm(avg_spending_usd ~ avg_stay_nights + origin + year,
                     data = visitors)
tidy(spending_model) |>
  mutate(across(where(is.numeric), ~ round(., 3)))
```

```{r workflow-step6}
# Step 6: Model fit
glance(spending_model) |>
  select(r.squared, adj.r.squared, p.value, AIC)
```

This entire analysis --- from descriptives to regression --- is captured in a
script that you can re-run at any time. If your data updates or a reviewer
requests changes, you modify the code and run it again. No clicking through
dialog boxes.

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Complete analysis workflow

Conduct a full analysis to answer this question: **Do satisfaction scores differ
between high-season (Q1 and Q4) and low-season (Q2 and Q3) quarters?**

Your workflow should include:

1. Create a new variable `season` that classifies Q1/Q4 as "High" and Q2/Q3 as
   "Low"
2. Compute descriptive statistics (mean and SD of `satisfaction_score`) for each
   season
3. Create a boxplot comparing satisfaction scores by season
4. Run an independent-samples t-test
5. Use `broom::tidy()` to extract the results into a clean table

:::::::::::::::::::::::: solution

## Solution

```{r challenge1-solution}
# Step 1: Create the season variable
visitors_season <- visitors |>
  mutate(season = if_else(quarter %in% c("Q1", "Q4"), "High", "Low"))

# Step 2: Descriptive statistics
visitors_season |>
  group_by(season) |>
  summarise(
    n = n(),
    mean_satisfaction = mean(satisfaction_score),
    sd_satisfaction = sd(satisfaction_score)
  )

# Step 3: Boxplot
ggplot(visitors_season, aes(x = season, y = satisfaction_score,
                             fill = season)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Satisfaction Scores: High vs. Low Season",
    x = "Season",
    y = "Satisfaction Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Step 4: T-test
season_t <- t.test(satisfaction_score ~ season, data = visitors_season)
season_t

# Step 5: Tidy results
tidy(season_t)
```

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Correlation and regression

Investigate the relationship between `avg_stay_nights` and `avg_spending_usd`:

1. Run `cor.test()` to get the Pearson correlation and p-value
2. Create a scatterplot with a linear trend line (`geom_smooth(method = "lm")`)
3. Fit a linear regression predicting `avg_spending_usd` from `avg_stay_nights`
4. Use `tidy()` and `glance()` to extract the results
5. Interpret: for each additional night of stay, how much does spending increase?

:::::::::::::::::::::::: solution

## Solution

```{r challenge2-solution}
# Step 1: Correlation
cor.test(visitors$avg_stay_nights, visitors$avg_spending_usd)

# Step 2: Scatterplot
ggplot(visitors, aes(x = avg_stay_nights, y = avg_spending_usd)) +
  geom_point(aes(color = origin), alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", color = "black", linewidth = 0.8) +
  labs(
    title = "Relationship Between Length of Stay and Spending",
    x = "Average Stay (Nights)",
    y = "Average Spending (USD)",
    color = "Origin"
  ) +
  theme_minimal()

# Step 3: Regression
stay_model <- lm(avg_spending_usd ~ avg_stay_nights, data = visitors)

# Step 4: Extract results
tidy(stay_model)
glance(stay_model)

# Step 5: Interpretation
# The coefficient for avg_stay_nights tells us:
# for each additional night of stay, average spending increases by
# approximately the "estimate" value in USD, holding all else constant.
```

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- Every SPSS statistical test has a direct R equivalent, usually in a single function call
- R output is more compact than SPSS --- `broom::tidy()` converts it to a clean table
- The workflow in R is: load data, run test, extract results, visualize --- all in a script

::::::::::::::::::::::::::::::::::::::::::::::::
