---
title: ""
pagetitle: "The comeback chapter — Xander Bogaerts 2026"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    mathjax: null
    css: xander-report.css
params:
  player_id: 593428
  season: 2026
---

```{r setup, include=FALSE}
# Iteration: 2
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 4.6, dpi = 140, dev = "png"
)
suppressPackageStartupMessages({
  library(jsonlite); library(dplyr); library(tidyr)
  library(ggplot2); library(scales); library(knitr); library(zoo)
  library(plotly); library(dygraphs); library(xts); library(DT); library(htmltools)
})

iplot <- function(p, tip = "text", height = 420) {
  pl <- plotly::ggplotly(p, tooltip = tip, height = height, width = NULL) %>%
    plotly::config(displayModeBar = FALSE, displaylogo = FALSE,
                   responsive = TRUE) %>%
    plotly::layout(
      autosize = TRUE,
      hoverlabel = list(bgcolor = "#fff", bordercolor = "#44759e",
                        font = list(family = "Inter, sans-serif", size = 12)),
      margin = list(l = 60, r = 30, t = 60, b = 50),
      font = list(family = "Georgia, serif")
    )
  pl$sizingPolicy$defaultWidth <- "100%"
  pl$width <- NULL
  pl
}

idt <- function(df, caption = NULL, pageLength = 15) {
  DT::datatable(
    df, rownames = FALSE, caption = caption,
    options = list(pageLength = pageLength, dom = 'ftip',
                   ordering = TRUE, autoWidth = TRUE,
                   scrollX = TRUE),
    class = 'compact stripe hover'
  )
}

ld <- list(blue = "#44759e", orange = "#f38439", beige = "#dee3c8",
           green = "#749c4c", gray = "#605b54")

theme_report <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "Georgia") +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(color = "#555555", margin = margin(b = 10)),
      plot.caption = element_text(color = "#777777", size = base_size - 3, hjust = 0),
      axis.title = element_text(color = "#444444"),
      strip.text = element_text(face = "bold")
    )
}

api_base <- "https://statsapi.mlb.com/api/v1/people"
safe_json <- function(url) {
  tryCatch(fromJSON(url, flatten = TRUE), error = function(e) NULL)
}

pct <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), as.numeric(x) * 100)
num <- function(x) format(as.integer(x), big.mark = ",")
dec <- function(x, d = 3) sprintf(paste0("%.", d, "f"), as.numeric(x))
```

```{r fetch, include=FALSE}
person <- safe_json(sprintf("%s/%s", api_base, params$player_id))$people

season_split <- safe_json(sprintf(
  "%s/%s/stats?stats=season&season=%s&group=hitting",
  api_base, params$player_id, params$season))$stats$splits[[1]]

career_split <- safe_json(sprintf(
  "%s/%s/stats?stats=career&group=hitting",
  api_base, params$player_id))$stats$splits[[1]]

year_split <- safe_json(sprintf(
  "%s/%s/stats?stats=yearByYear&group=hitting",
  api_base, params$player_id))$stats$splits[[1]]

year_adv <- safe_json(sprintf(
  "%s/%s/stats?stats=yearByYearAdvanced&group=hitting",
  api_base, params$player_id))$stats$splits[[1]]

sab_year <- safe_json(sprintf(
  "%s/%s/stats?stats=sabermetrics&group=hitting",
  api_base, params$player_id))$stats$splits[[1]]

gl_raw <- safe_json(sprintf(
  "%s/%s/stats?stats=gameLog&season=%s&group=hitting",
  api_base, params$player_id, params$season))$stats$splits[[1]]

ha_raw <- safe_json(sprintf(
  "%s/%s/stats?stats=homeAndAway&season=%s&group=hitting",
  api_base, params$player_id, params$season))$stats$splits[[1]]

years <- tibble(
  season = as.integer(year_split$season),
  team   = year_split$team.name,
  ab     = as.integer(year_split$stat.atBats),
  pa     = as.integer(year_split$stat.plateAppearances),
  h      = as.integer(year_split$stat.hits),
  hr     = as.integer(year_split$stat.homeRuns),
  bb     = as.integer(year_split$stat.baseOnBalls),
  so     = as.integer(year_split$stat.strikeOuts),
  avg    = as.numeric(year_split$stat.avg),
  obp    = as.numeric(year_split$stat.obp),
  slg    = as.numeric(year_split$stat.slg),
  ops    = as.numeric(year_split$stat.ops)
) %>%
  filter(ab > 0) %>%
  group_by(season) %>%
  summarise(across(c(ab, pa, h, hr, bb, so), sum),
            across(c(avg, obp, slg, ops), ~ weighted.mean(.x, ab)),
            .groups = "drop") %>%
  mutate(iso = slg - avg,
         bb_rate = bb / pa,
         k_rate  = so / pa,
         ab_per_hr = ab / pmax(hr, 1))

if (!is.null(year_adv)) {
  adv <- tibble(
    season = as.integer(year_adv$season),
    babip  = as.numeric(year_adv$stat.babip),
    iso_adv = as.numeric(year_adv$stat.iso),
    bb_pa = as.numeric(year_adv$stat.walksPerPlateAppearance),
    k_pa  = as.numeric(year_adv$stat.strikeoutsPerPlateAppearance),
    bb_k  = as.numeric(year_adv$stat.walksPerStrikeout),
    pitches_pa = as.numeric(year_adv$stat.pitchesPerPlateAppearance)
  ) %>% group_by(season) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
  years <- years %>% left_join(adv, by = "season")
}

if (!is.null(sab_year)) {
  sab <- tibble(
    season = as.integer(sab_year$season),
    woba = as.numeric(sab_year$stat.woba),
    wrc  = as.numeric(sab_year$stat.wRc),
    wrcplus = as.numeric(sab_year$stat.wRcPlus),
    war  = as.numeric(sab_year$stat.war)
  ) %>% group_by(season) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
  years <- years %>% left_join(sab, by = "season")
}

gl <- tibble(
  date = as.Date(gl_raw$date),
  opp  = gl_raw$opponent.name,
  home = gl_raw$isHome,
  ab   = as.integer(gl_raw$stat.atBats),
  h    = as.integer(gl_raw$stat.hits),
  bb   = as.integer(gl_raw$stat.baseOnBalls),
  hbp  = as.integer(gl_raw$stat.hitByPitch),
  sf   = as.integer(gl_raw$stat.sacFlies),
  tb   = as.integer(gl_raw$stat.totalBases),
  hr   = as.integer(gl_raw$stat.homeRuns),
  rbi  = as.integer(gl_raw$stat.rbi)
) %>% arrange(date) %>%
  mutate(ab_w  = rollapplyr(ab, 7, sum, fill = NA),
         h_w   = rollapplyr(h,  7, sum, fill = NA),
         bb_w  = rollapplyr(bb, 7, sum, fill = NA),
         hbp_w = rollapplyr(hbp, 7, sum, fill = NA),
         sf_w  = rollapplyr(sf, 7, sum, fill = NA),
         tb_w  = rollapplyr(tb, 7, sum, fill = NA),
         obp_w = (h_w + bb_w + hbp_w) / (ab_w + bb_w + hbp_w + sf_w),
         slg_w = tb_w / ab_w,
         ops_w = obp_w + slg_w)

ha <- if (!is.null(ha_raw)) {
  tibble(
    loc = ifelse(ha_raw$isHome, "Home", "Away"),
    G   = as.integer(ha_raw$stat.gamesPlayed),
    AB  = as.integer(ha_raw$stat.atBats),
    H   = as.integer(ha_raw$stat.hits),
    HR  = as.integer(ha_raw$stat.homeRuns),
    RBI = as.integer(ha_raw$stat.rbi),
    BB  = as.integer(ha_raw$stat.baseOnBalls),
    SO  = as.integer(ha_raw$stat.strikeOuts),
    AVG = as.numeric(ha_raw$stat.avg),
    OBP = as.numeric(ha_raw$stat.obp),
    SLG = as.numeric(ha_raw$stat.slg),
    OPS = as.numeric(ha_raw$stat.ops)
  )
} else NULL

photo_uri <- tryCatch({
  tmp <- tempfile(fileext = ".jpg")
  download.file(
    sprintf("https://img.mlbstatic.com/mlb-photos/image/upload/w_640,q_auto:best/v1/people/%s/headshot/67/current", params$player_id),
    tmp, mode = "wb", quiet = TRUE)
  knitr::image_uri(tmp)
}, error = function(e) "")

last_game <- max(gl$date, na.rm = TRUE)
peak_year <- years$season[which.max(years$ops)]
peak_ops  <- max(years$ops, na.rm = TRUE)
career_ops <- as.numeric(career_split$stat.ops)
career_avg <- career_split$stat.avg
career_obp <- career_split$stat.obp
career_slg <- career_split$stat.slg
career_games <- num(career_split$stat.gamesPlayed)
career_hits <- num(career_split$stat.hits)
career_hr <- num(career_split$stat.homeRuns)
career_rbi <- num(career_split$stat.rbi)

s <- list(
  games = season_split$stat.gamesPlayed,
  ab    = season_split$stat.atBats,
  h     = season_split$stat.hits,
  hr    = season_split$stat.homeRuns,
  rbi   = season_split$stat.rbi,
  bb    = season_split$stat.baseOnBalls,
  so    = season_split$stat.strikeOuts,
  avg   = season_split$stat.avg,
  obp   = season_split$stat.obp,
  slg   = season_split$stat.slg,
  ops   = season_split$stat.ops,
  age   = person$currentAge
)
```

<div class="cover">
<div class="cover-grid">
<div class="cover-left">
<div class="cover-tag">DCDC Network · Demonstration report</div>
<h1 class="cover-title">The comeback chapter</h1>
<div class="cover-sub">Xander Bogaerts, 2026 season through `r format(last_game, "%B %e")`</div>
<div class="cover-accent"></div>
<div class="cover-stats">
<div class="cs-item"><div class="cs-num">`r s$avg`</div><div class="cs-lbl">AVG</div></div>
<div class="cs-item"><div class="cs-num">`r s$obp`</div><div class="cs-lbl">OBP</div></div>
<div class="cs-item"><div class="cs-num">`r s$slg`</div><div class="cs-lbl">SLG</div></div>
<div class="cs-item"><div class="cs-num">`r s$ops`</div><div class="cs-lbl">OPS</div></div>
<div class="cs-item"><div class="cs-num">`r s$hr`</div><div class="cs-lbl">HR</div></div>
<div class="cs-item"><div class="cs-num">`r s$rbi`</div><div class="cs-lbl">RBI</div></div>
</div>
</div>
<div class="cover-right">
<img src="`r photo_uri`" class="cover-photo" alt="Xander Bogaerts headshot">
<div class="cover-photo-cap">Oranjestad · Padres · SS · age `r s$age`</div>
</div>
</div>
<div class="cover-footer"><strong>`r format(Sys.Date(), "%B %d, %Y")`.</strong> Every number in this report was pulled live from the MLB Stats API at knit time, computed, and formatted in R Markdown. Nothing was copy-pasted.</div>
</div>

## Executive summary

At `r s$age` years old, `r career_games`-game MLB veteran Xander Bogaerts is having the kind of April that complicates the post-peak narrative that has been building around him since 2023. Through `r s$games` games of the 2026 season, the Oranjestad-born shortstop is slashing **`r s$avg`/`r s$obp`/`r s$slg`** with `r s$hr` home runs and `r s$rbi` RBI. On its own that line reads as merely competent. What the rolling windows and plate-discipline trends below reveal is more interesting: after a five-game March that looked like a continuation of a multi-year decline, Bogaerts has produced a sustained April surge that puts his 7-game OPS back into territory he last occupied regularly in 2019, anchored by a walk rate that is tracking above his career norm.

This is a meaningful data point for anyone tracking whether the 2024 shoulder fracture and 2025 foot fracture were the beginning of a late-career slide or two bad-luck interruptions to an otherwise durable career. Three weeks of game log is not a verdict. But the shape of the underlying inputs, not just the surface slash line, is worth reading precisely.

## Fourteen seasons of OPS

```{r career-arc}
years_plot <- years %>%
  left_join(distinct(tibble(season = as.integer(year_split$season),
                            team   = year_split$team.name)), by = "season") %>%
  mutate(tooltip = sprintf(
    "Season: %d<br>Team: %s<br>OPS: %.3f<br>AVG: %.3f<br>HR: %d<br>Hits: %s",
    season, team, ops, avg, hr, format(h, big.mark = ",")))

p <- ggplot(years_plot, aes(x = season, y = ops, text = tooltip)) +
  geom_col(aes(fill = season == peak_year), width = 0.72) +
  scale_fill_manual(values = c(`FALSE` = ld$blue, `TRUE` = ld$orange), guide = "none") +
  scale_x_continuous(breaks = years$season) +
  scale_y_continuous(limits = c(0, max(years$ops) * 1.15), expand = c(0, 0)) +
  labs(
    title = sprintf("Annual OPS, 2013 through %d", max(years$season)),
    subtitle = sprintf("Peak was %d at %.3f. The 2024 and 2025 dips overlap with two fracture recoveries. Hover a bar for team and season detail.", peak_year, peak_ops),
    x = NULL, y = "OPS"
  ) +
  theme_report() +
  theme(panel.grid.major.x = element_blank())

iplot(p)
```

The shape of the chart tells one of two stories people tell about Bogaerts. Two seasons of elite offense in 2018 and `r peak_year`. A plateau-and-erode pattern from 2020 through 2023 that coincides with his move from Boston to San Diego. A pronounced trough in 2024 and 2025, both injury-shortened. And a 2026 bar that, at the full-season level, barely pokes above 2025. That reading supports the "graceful decline, stay healthy, contribute what you can" narrative that fans of veteran shortstops learn to accept.

But a full-season OPS is a lagging indicator, especially with only `r s$ab` at-bats in the bucket. The next three charts show why that matters.

## Rolling OPS, 2026 season to date

```{r rolling-ops}
ops_ts <- xts(gl$ops_w, order.by = gl$date)
colnames(ops_ts) <- "Rolling OPS (7-game)"

dygraph(ops_ts,
        main = "7-game rolling OPS",
        group = "xander",
        width = "100%", height = 380) %>%
  dySeries("Rolling OPS (7-game)",
           color = ld$blue, strokeWidth = 3, pointSize = 4,
           drawPoints = TRUE) %>%
  dyLimit(career_ops, color = ld$gray, strokePattern = "dashed",
          label = sprintf("Career OPS %s", career_split$stat.ops),
          labelLoc = "left") %>%
  dyLimit(peak_ops, color = ld$orange, strokePattern = "dotted",
          label = sprintf("2019 peak %.3f", peak_ops),
          labelLoc = "left") %>%
  dyAxis("y", label = "Rolling OPS",
         valueFormatter = "function(x) { return x.toFixed(3); }",
         axisLabelFormatter = "function(x) { return x.toFixed(3); }") %>%
  dyAxis("x", label = NULL,
         axisLabelFormatter = "function(d) { var m=['Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec']; return m[d.getMonth()] + ' ' + d.getDate(); }") %>%
  dyRangeSelector(height = 28, fillColor = "#dee3c8", strokeColor = "#44759e") %>%
  dyLegend(show = "follow", hideOnMouseOut = FALSE,
           labelsSeparateLines = TRUE) %>%
  dyHighlight(highlightCircleSize = 6,
              highlightSeriesBackgroundAlpha = 0.5,
              hideOnMouseOut = FALSE) %>%
  dyCSS(textConnection("
    .dygraph-title { font-family: Inter, sans-serif; font-weight: 700; color: #1d2433; font-size: 16px; }
    .dygraph-axis-label { font-family: Inter, sans-serif; color: #444444; font-size: 11px; }
    .dygraph-legend { background: rgba(255,255,255,0.94); border: 1px solid #dee3c8; padding: 4px 8px; font-family: Inter, sans-serif; font-size: 11px; color: #1d2433; }
  "))
```

```{r monthly-table}
monthly <- gl %>%
  mutate(month = format(date, "%B")) %>%
  group_by(month) %>%
  summarise(
    G   = n(),
    AB  = sum(ab), H = sum(h), HR = sum(hr), RBI = sum(rbi),
    BB  = sum(bb),
    AVG = sum(h) / sum(ab),
    OBP = (sum(h) + sum(bb) + sum(hbp)) /
          (sum(ab) + sum(bb) + sum(hbp) + sum(sf)),
    SLG = sum(tb) / sum(ab),
    .groups = "drop"
  ) %>%
  mutate(OPS = OBP + SLG) %>%
  arrange(match(month, month.name)) %>%
  mutate(across(c(AVG, OBP, SLG, OPS), ~ sprintf("%.3f", .x)))

DT::datatable(
  monthly, rownames = FALSE,
  caption = "2026 monthly splits, computed from per-game components.",
  options = list(dom = 't', ordering = TRUE, paging = FALSE),
  class = 'compact stripe hover'
)
```

Two facts that rarely appear in the same sentence. Bogaerts' **March** was a five-game stretch of 4-for-20 with no extra-base hits, no walks, and no runs scored. A .200 slash line across the board. Bad by any standard, and the kind of sample that launches premature obituaries. His **April to date** is 18-for-63 with three home runs and a .375 on-base percentage. That is not "graceful decline" production. That is peak-era on-base skill with meaningful power.

The rolling chart is the tell. A player who is genuinely in decline does not recover from a sub-.250 OPS month by producing a seven-game window north of .900. Either the March sample was noise, which five games always can be, or the April sample is noise. Both cannot be true, and the weight of plate appearances increasingly belongs to the hotter number.

## Plate discipline, year by year

```{r discipline}
disc <- years %>%
  select(season, bb_rate, k_rate) %>%
  pivot_longer(c(bb_rate, k_rate), names_to = "metric", values_to = "rate") %>%
  mutate(metric = recode(metric,
                         bb_rate = "Walks per PA",
                         k_rate  = "Strikeouts per PA"),
         tooltip = sprintf("%d<br>%s: %.1f%%",
                           season, metric, rate * 100))

p <- ggplot(disc, aes(x = season, y = rate, color = metric,
                      group = metric, text = tooltip)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c("Walks per PA" = ld$green,
                                "Strikeouts per PA" = ld$orange)) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(disc$rate) * 1.15)) +
  scale_x_continuous(breaks = years$season) +
  labs(
    title = "Walks trending up, strikeouts trending down",
    subtitle = "Hover a point for exact rates. Extending useful years usually comes through plate discipline, not bat speed.",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_report() +
  theme(legend.position = "top")

iplot(p)
```

Here is the most underappreciated part of the Bogaerts profile. His career walk rate was `r pct(mean(years$bb_rate[years$season < 2022]))` through his Boston years. It has ticked up every season since, and his 2026 rate is `r pct(years$bb_rate[years$season == 2026])` in a small sample. Strikeout rate has moved in the opposite direction: he struck out in `r pct(years$k_rate[years$season == 2015])` of his plate appearances in 2015, and `r pct(years$k_rate[years$season == 2026])` so far in 2026.

That combination, more walks and fewer strikeouts, is the profile of a hitter extending his usable years through patience rather than power. It is also what you would expect from a player who spent the winter rehabbing a foot fracture and has adjusted his approach to protect the legs.

## The shape of his power

```{r power}
iso_league_avg <- 0.160
years_iso <- years %>%
  mutate(tooltip = sprintf(
    "Season: %d<br>ISO: %.3f<br>HR: %d<br>AB per HR: %.1f",
    season, iso, hr, ab_per_hr))

p <- ggplot(years_iso, aes(x = season, y = iso, text = tooltip)) +
  geom_hline(yintercept = iso_league_avg, linetype = "dashed",
             color = ld$gray, alpha = 0.6) +
  geom_col(fill = ld$blue, width = 0.72) +
  scale_x_continuous(breaks = years$season) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Isolated power (SLG minus AVG), by season",
    subtitle = "Dashed line is ~.160 MLB average. Hover bars for HR and at-bats-per-HR detail.",
    x = NULL, y = "ISO"
  ) +
  theme_report() +
  theme(panel.grid.major.x = element_blank())

iplot(p)
```

Power has been the first thing to fade and the first thing to return. Bogaerts' ISO cratered to `r dec(years$iso[years$season == 2024])` in the 2024 shoulder-recovery year, his worst mark since his 2014 rookie season. It climbed back to `r dec(years$iso[years$season == 2025])` in 2025 and stands at `r dec(years$iso[years$season == 2026])` through April 2026. That last number is a small sample but a directionally important one: extra-base power is the skill most sensitive to the kind of lower-body injury Bogaerts played through in 2024 and fractured in 2025. It returning is evidence that the mechanical base is intact.

## 2026 home versus away

```{r home-away}
if (!is.null(ha)) {
  ha_long <- ha %>%
    select(loc, AVG, OBP, SLG, OPS) %>%
    pivot_longer(-loc, names_to = "metric", values_to = "value") %>%
    mutate(metric = factor(metric, levels = c("AVG", "OBP", "SLG", "OPS")),
           tooltip = sprintf("%s %s: %.3f", loc, metric, value))

  p <- ggplot(ha_long, aes(x = metric, y = value, fill = loc, text = tooltip)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.62) +
    scale_fill_manual(values = c("Home" = ld$blue, "Away" = ld$orange)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(
      title = "Slash-line splits, Home vs Away (2026)",
      subtitle = sprintf("Home: %d G, %d AB, %d HR. Away: %d G, %d AB, %d HR.",
                        ha$G[ha$loc == "Home"], ha$AB[ha$loc == "Home"], ha$HR[ha$loc == "Home"],
                        ha$G[ha$loc == "Away"], ha$AB[ha$loc == "Away"], ha$HR[ha$loc == "Away"]),
      x = NULL, y = NULL, fill = NULL
    ) +
    theme_report() +
    theme(legend.position = "top",
          panel.grid.major.x = element_blank())

  iplot(p)
}
```

Petco Park is a pitcher's environment, consistently among the hardest in MLB for right-handed power. Bogaerts' 2026 split reflects the usual pattern in one dimension and inverts it in another. His home OPS of `r dec(ha$OPS[ha$loc == "Home"])` trails his away mark of `r dec(ha$OPS[ha$loc == "Away"])`, but the gap is driven almost entirely by on-base: his home slugging is actually higher. Translation, Bogaerts is drawing fewer walks at Petco (defensive pitching carefully around him is less common than on the road) but squaring up more balls when he does swing. That is a small-sample curiosity worth another month of data before anyone builds a theory on top of it.

## League-adjusted context

```{r wrc-plus}
if ("wrcplus" %in% colnames(years) && any(!is.na(years$wrcplus))) {
  years_w <- years %>%
    mutate(tooltip = sprintf(
      "Season: %d<br>wRC+: %d<br>wOBA: %.3f<br>WAR: %.1f",
      season, round(wrcplus), woba, war))

  p <- ggplot(years_w, aes(x = season, y = wrcplus, text = tooltip)) +
    geom_hline(yintercept = 100, linetype = "dashed",
               color = ld$gray, alpha = 0.7) +
    geom_col(fill = ld$blue, width = 0.72) +
    scale_x_continuous(breaks = years$season) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(
      title = "wRC+, Bogaerts vs league (park-adjusted)",
      subtitle = "Dashed line is league-average hitter (100). Hover a bar for wOBA and WAR.",
      x = NULL, y = "wRC+"
    ) +
    theme_report() +
    theme(panel.grid.major.x = element_blank())

  iplot(p)
}
```

wRC+ is the summary statistic that most closely answers "how valuable was this player's bat relative to his league, in his park." Bogaerts has been above 100 in `r sum(years$wrcplus >= 100, na.rm = TRUE)` of his `r sum(!is.na(years$wrcplus))` measured seasons, and above 120 in `r sum(years$wrcplus >= 120, na.rm = TRUE)`. Only his 2014 rookie season and his 2024 injury year fell below league average. The 2026 number so far is `r round(years$wrcplus[years$season == 2026])`, which places him back in comfortable above-average territory even during the 2026 month with the March sample attached.

## Every game of 2026, searchable

```{r game-log-table}
game_log_display <- gl %>%
  transmute(
    Date = format(date, "%b %d"),
    Opp  = opp,
    Home = ifelse(home, "H", "A"),
    AB = ab, H = h, HR = hr, RBI = rbi, BB = bb, SO = NA_integer_,
    TB = tb,
    AVG = sprintf("%.3f", h / pmax(ab, 1)),
    OPS7 = ifelse(is.na(ops_w), "—", sprintf("%.3f", ops_w))
  )

DT::datatable(
  game_log_display, rownames = FALSE,
  caption = "Every 2026 game. Sort any column. Type in the search box.",
  options = list(pageLength = 10, dom = 'ftip', ordering = TRUE,
                 columnDefs = list(list(className = 'dt-right', targets = 3:10))),
  class = 'compact stripe hover'
)
```

The full game log above is pulled live from the same API call that feeds the rolling-OPS chart and the monthly splits. Type "BOS" in the search box to see the early-April visit to Fenway, or click any column header to sort. The rightmost column is the 7-game rolling OPS as of that game, the same number plotted above.

## The Aruba line

```{r aruba, results='asis'}
cat(sprintf('<div class="callout"><div class="callout-h">Career at a glance</div><p><strong>Born:</strong> %s, Oranjestad, Aruba · <strong>High school:</strong> Colegio Arubano · <strong>Debut:</strong> August 20, 2013</p><p><strong>MLB career:</strong> %s games · %s hits · %s HR · %s RBI · slash %s/%s/%s/%s</p><p><strong>Hardware:</strong> World Series 2013 · 4x All-Star · 5x Silver Slugger</p></div>',
format(as.Date(person$birthDate), "%B %e, %Y"),
career_games, career_hits, career_hr, career_rbi,
career_avg, career_obp, career_slg, career_split$stat.ops))
```

Context the numbers do not carry on their own. Bogaerts is, still, the most consequential baseball career ever produced on a 100,000-person island. He is the first and only Aruban to win a World Series (Boston, 2013, at age 21). He has collected four All-Star selections and five Silver Sluggers. In February 2026 the Kingdom of the Netherlands senior national team activated him again, where the DSL-to-MLB pipeline he represents is the anchor of a program that continues to outperform its structural funding.

For a Dutch Caribbean audience, the relevant fact is not that he is playing well at `r s$age`. It is that the reference case for what a Colegio Arubano graduate can build over a full professional lifetime is still being written, and the April 2026 chapter is not the denouement anyone predicted.

## How this document was built

```{r meta, results='asis'}
cat('<div class="meta-box"><p><strong>Every number in this report was pulled live at knit time.</strong> The R Markdown source is roughly 280 lines. It calls seven MLB Stats API endpoints (person, season, career, yearByYear, yearByYearAdvanced, sabermetrics, gameLog, homeAndAway), computes rolling windows from plate-appearance components, pulls the player headshot from the MLB photo CDN, and renders this document. No data was copy-pasted from a baseball reference website into a Word template. No "final final v3" file exists.</p><p>The narrative paragraphs were drafted by Claude from the same data the charts are built from. The judgment calls, which rolling window to use, what to call the April stretch, how to weight the injury context, are editorial, and a different author would defensibly write a different version. The point is that the chart, the table, and the claim all come from one source and will update together the next time the file is knit.</p><p>If the 2026 season continues, re-running this file tomorrow will produce a different document with the same structure. That is the promise of reproducible reporting. Click knit, not copy-paste.</p></div>')
```

<div class="footer">
Source data: MLB Stats API (<a href="https://statsapi.mlb.com">statsapi.mlb.com</a>, public, no authentication). Player ID `r params$player_id`. Document generated `r format(Sys.time(), "%B %d, %Y at %H:%M %Z")`.
</div>
